Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the transition model $\hat{z}_{t+1} = d_{\phi}(e_{\phi}(z_{t}), a_{t})$
and the next state predicted by the momentum encoder $\bar{z}_{t+1} = e_{\bar{\phi}}(z_{t+1})$.
The prediction $d_{\phi}(e_{\phi}(z_{t}), a_{t})$ uses the online encoder, which is being trained jointly with the transition model.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder, which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
% As such, gradient to not flow through the target network
The target network update rate is denoted $\tau$.

@@

FSQ has the following hyperparameters: we must specify the number of channels $c$ and the number of levels per channel
$\mathcal{L} = [L_{1},\ldots,L_{d}]$.
\cref{tab:fsq-levels} shows the recommended number of channels and number of levels per channel to obtain codebooks of
different sizes.
FSQ, relies on a simple, fixed grid partition in a much
lower-dimensional space.
In practice, we found codebooks of size $|\mathcal{C}|=2^{6}$ sufficient for most environments in the DeepMind Control suite.
However, for the Dog Walk environment we used a codebook of size $|\mathcal{C}|=2^{9}$.

@@

\textbf{Policy and value function learning}
We learn the policy and value function using TD3 \citep{fujimotoAddressingFunctionApproximation2018}.
However, we follow \citet{yaratsMasteringVisualContinuous2021,zhaoSimplifiedTemporalConsistency2023}
and augment the loss with n-step returns.
The only difference to TD3 is that instead of using the original observations $o_{t}$, we map them through the
encoder $z_{t} = e_{\bar{\theta}}(o_{t})$ and learn the value function and policy in the latent space $z_{t}$.
The value function is updated by minimizing the following objective:
\begin{align} \label{eq:value-loss}
  \mathcal{L}_{Q} &= \E_{\tau \sim \mathcal{D}} \left[ (Q_{\psi_{k}}(e_{\bar{\theta}}(s_{t}), a_{t}) - y)^{2}  \right], \quad  \forall k \in 1, 2 \\
  y &= \sum_{n=0}^{N-1} \gamma^{n} r_{t+n} + \gamma^{n} \min_{k \in \{1,2\}} Q_{\bar{\psi}_{k}}(e_{\bar{\theta}}(s_{t+n}), a_{t+n}), \nonumber
\end{align}
where we use policy smoothing by adding clipped Gaussian noise $\epsilon_{t+n} \sim \mathcal{N} (0,\sigma^{2})$ to the
action $a_{t+n} = \pi_{\bar{\eta}}(z_{t+n}) + \epsilon_{t+n}$.
Note that we use the target encoder to get the latent states in both the prediction and the target.
We then use the target value functions $Q_{\bar{\psi}_{k}}$ and the target policy $\pi_{\bar{\eta}}$ to calculate the TD target.
Following TD3, we learn the actor's parameters by minimizing
%
\begin{align} \label{eq:policy-loss}
 \mathcal{L}_{\pi} = - \E_{(o_{t}, a_{t}) \sim \mathcal{D}} \left[ \min_{k\in\{1,2\}} Q_{\psi_{k}}(e_{\bar{\theta}}(o_{t}), a_{t}) \right].
\end{align}
%
That is, we maximize the Q-value using the clipped double Q learning trick to combat overestimation in Q learning.

@@

\textbf{Reward predict is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learning loss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $\tilde{r}_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\hat{z}_{t+h} &= d_{\phi}(e_{\theta}(s_{t}), a_{t}) \\
		\tilde{z}_{t+h} &= e_{\bar{\theta}}(s_{t+1}) \\
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right )
	\end{align*}
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to not suffering from latent collapse in this specific environment. Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes learning the representation learning task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

>>>>

\textbf{Representation learning}
Our representation learning is based solely on the latent-state consistency loss,
%
\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
% &= \E_{\tau \sim \mathcal{D}}
% \left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
&= \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(\hat{z}_{t}, a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2},
\end{align}
%
which minimizes the distance between the next state predicted by the transition model $\hat{z}_{t+1} = d_{\phi}(\hat{z}_{t}, a_{t})$
and the next state predicted by the momentum encoder $\bar{z}_{t+1} = e_{\bar{\phi}}(s_{t+1})$.
The latent states are obtained with multi-step predictions in the latent space $\hat{z}_{t+1} = d(\hat{z}_{t}, a_{t})$
with $\hat{z}_{0} = e_{\theta}(s_{0})$.
The initial mapping to the latent space $\hat{z}_{0} = e_{\theta}(s_{0})$ uses the online encoder which
is being trained jointly with the transition model $d_{\phi}(\hat{z_{t}}, a_{t})$.
The target $e_{\bar{\phi}}(z_{t+1})$ is calculated with the momentum encoder which uses an exponential moving average (EMA)
of the encoder's weights $\bar{\theta} \leftarrow \tau \bar{\theta} + (1-\tau)\theta$.
% As such, gradient to not flow through the target network
The target network update rate is denoted $\tau$.

@@

FSQ has the following hyperparameters: we must specify the number of channels $c$ and the number of levels per channel
$\mathcal{L} = [L_{1},\ldots,L_{d}]$.
\cref{tab:fsq-levels} shows the recommended number of channels and number of levels per channel to obtain codebooks of different sizes \citep{mentzerFiniteScalarQuantization2023}.
FSQ, relies on a simple, fixed grid partition in a much
lower-dimensional space.
In practice, we found codebooks of size $|\mathcal{C}|=2^{6}$ sufficient for most environments in the DeepMind Control suite.
However, for the Dog Walk environment we used a codebook of size $|\mathcal{C}|=2^{9}$.

@@

\textbf{Policy and value function learning}
We learn the policy and value function using TD3 \citep{fujimotoAddressingFunctionApproximation2018}.
However, we follow \citet{yaratsMasteringVisualContinuous2021,zhaoSimplifiedTemporalConsistency2023}
and augment the loss with n-step returns.
The only difference to TD3 is that instead of using the original observations $o_{t}$, we map them through the
target encoder $z_{t} = e_{\bar{\theta}}(o_{t})$ and learn the value function and policy in the latent space $z_{t}$.
The value function is updated by minimizing the following objective:
\begin{align} \label{eq:value-loss}
  \mathcal{L}_{Q} &= \E_{\tau \sim \mathcal{D}} \left[ (Q_{\psi_{k}}(e_{\bar{\theta}}(s_{t}), a_{t}) - y)^{2}  \right], \quad  \forall k \in 1, 2 \\
  y &= \sum_{n=0}^{N-1} \gamma^{n} r_{t+n} + \gamma^{n} \min_{k \in \{1,2\}} Q_{\bar{\psi}_{k}}(e_{\bar{\theta}}(s_{t+n}), a_{t+n}), \nonumber
\end{align}
where we use policy smoothing by adding clipped Gaussian noise $\epsilon_{t+n} \sim \mathcal{N} (0,\sigma^{2})$ to the
action $a_{t+n} = \pi_{\bar{\eta}}(z_{t+n}) + \epsilon_{t+n}$.
Note that we use the target encoder to get the latent states in both the prediction and the target.
We then use the target value functions $Q_{\bar{\psi}_{k}}$ and the target policy $\pi_{\bar{\eta}}$ to calculate the TD target.
Following TD3, we learn the actor's parameters by minimizing
%
\begin{align} \label{eq:policy-loss}
 \mathcal{L}_{\pi} = - \E_{(o_{t}, a_{t}) \sim \mathcal{D}} \left[ \min_{k\in\{1,2\}} Q_{\psi_{k}}(e_{\bar{\theta}}(o_{t}), a_{t}) \right].
\end{align}
%
That is, we maximize the Q-value using the clipped double Q learning trick to combat overestimation in Q learning.

@@

\textbf{Reward predict is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learning loss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\hat{z}_{t+h} &= d_{\phi}(\hat{z}_{t}, a_{t}) \\
		\tilde{z}_{t+h} &= e_{\bar{\theta}}(s_{t+1}) \\
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right )
	\end{align*}
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to not suffering from rank collapse in this specific environment. Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.
